---
title: "Code1"
output: pdf_document
date: "2025-05-06"
---
## Settings
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
```
## Reading data frames
```{r}
library(foreign)
library(tidyverse)
library(dplyr)
wave1 <- covid_19_wave1_survey_cls
wave2 <- covid_19_wave2_survey_cls
wave3 <- covid_19_wave3_survey_cls
```
## Exploring dataset, picking the data going to be used in later analysis
```{r}
wave1 <- wave1 %>%
  select(NCDSID, CW1_COVID19, CW1_COVID_HOSPAD, CW1_GHQ, CW1_HHNUM, CW1_MALAISE_1, CW1_MALAISE_2, CW1_MALAISE_3, CW1_MALAISE_4, CW1_MALAISE_5, CW1_MALAISE_6, CW1_MALAISE_7, CW1_MALAISE_8, CW1_MALAISE_9, CW1_GAD2PHQ2_1, CW1_GAD2PHQ2_2, CW1_GAD2PHQ2_3, CW1_GAD2PHQ2_4)
wave2 <- wave2 %>%
  select(NCDSID, CW2_COVID19, CW2_COVID_HOSPAD, CW2_GHQ, CW2_HHNUM, CW2_MALAISE_1, CW2_MALAISE_2, CW2_MALAISE_3, CW2_MALAISE_4, CW2_MALAISE_5, CW2_MALAISE_6, CW2_MALAISE_7, CW2_MALAISE_8, CW2_MALAISE_9, CW2_GAD2PHQ2_1, CW2_GAD2PHQ2_2, CW2_GAD2PHQ2_3, CW2_GAD2PHQ2_4)
wave3 <- wave3 %>%
  select(NCDSID, CW3_COVID19, CW3_COVID_HOSPAD, CW3_GHQ, CW3_HHNUM, CW3_MALAISE_1, CW3_MALAISE_2, CW3_MALAISE_3, CW3_MALAISE_4, CW3_MALAISE_5, CW3_MALAISE_6, CW3_MALAISE_7, CW3_MALAISE_8, CW3_MALAISE_9, CW3_GAD2PHQ2_1, CW3_GAD2PHQ2_2, CW3_GAD2PHQ2_3, CW3_GAD2PHQ2_4)
```
## Data cleaning, maybe we should filter out everyone who does not have a participant ID?
```{r}
wave1 <- wave1 %>%
  filter(NCDSID != "")
wave2 <- wave2 %>%
  filter(NCDSID != "")
wave3 <- wave3 %>%
  filter(NCDSID != "")
```
## I filtered all participants who participated in all three waves of data collections. Now we can look at longitudinal effects.
```{r}
common_ids <- reduce(list(wave1$NCDSID, wave2$NCDSID, wave3$NCDSID), intersect)
wave1_com <- wave1 %>%
  filter(NCDSID %in% common_ids)
wave2_com <- wave2 %>%
  filter(NCDSID %in% common_ids)
wave3_com <- wave3 %>%
  filter(NCDSID %in% common_ids)
```
## Specifying waves, changing variable names, pivot longer.
```{r}
wave1_com <- wave1_com %>%
  mutate(wave = 1)
wave2_com <- wave2_com %>%
  mutate(wave =2)
wave3_com <- wave3_com %>%
  mutate(wave = 3)

wave1_com <- wave1_com %>%
  rename(COVID19 = CW1_COVID19, COVID_HOSPAD = CW1_COVID_HOSPAD, GHQ = CW1_GHQ, HHNUM = CW1_HHNUM, MALAISE_1 = CW1_MALAISE_1, MALAISE_2 = CW1_MALAISE_2, MALAISE_3 = CW1_MALAISE_3, MALAISE_4 = CW1_MALAISE_4, MALAISE_5 = CW1_MALAISE_5, MALAISE_6 = CW1_MALAISE_6, MALAISE_7 = CW1_MALAISE_7, MALAISE_8 = CW1_MALAISE_8, MALAISE_9 = CW1_MALAISE_9, GAD2PHQ2_1 = CW1_GAD2PHQ2_1, GAD2PHQ2_2 = CW1_GAD2PHQ2_2, GAD2PHQ2_3 = CW1_GAD2PHQ2_3, GAD2PHQ2_4 = CW1_GAD2PHQ2_4)
wave2_com <- wave2_com %>%
  rename(COVID19 = CW2_COVID19, COVID_HOSPAD = CW2_COVID_HOSPAD, GHQ = CW2_GHQ, HHNUM = CW2_HHNUM, MALAISE_1 = CW2_MALAISE_1, MALAISE_2 = CW2_MALAISE_2, MALAISE_3 = CW2_MALAISE_3, MALAISE_4 = CW2_MALAISE_4, MALAISE_5 = CW2_MALAISE_5, MALAISE_6 = CW2_MALAISE_6, MALAISE_7 = CW2_MALAISE_7, MALAISE_8 = CW2_MALAISE_8, MALAISE_9 = CW2_MALAISE_9, GAD2PHQ2_1 = CW2_GAD2PHQ2_1, GAD2PHQ2_2 = CW2_GAD2PHQ2_2, GAD2PHQ2_3 = CW2_GAD2PHQ2_3, GAD2PHQ2_4 = CW2_GAD2PHQ2_4)
wave3_com <- wave3_com %>%
  rename(COVID19 = CW3_COVID19, COVID_HOSPAD = CW3_COVID_HOSPAD, GHQ = CW3_GHQ, HHNUM = CW3_HHNUM, MALAISE_1 = CW3_MALAISE_1, MALAISE_2 = CW3_MALAISE_2, MALAISE_3 = CW3_MALAISE_3, MALAISE_4 = CW3_MALAISE_4, MALAISE_5 = CW3_MALAISE_5, MALAISE_6 = CW3_MALAISE_6, MALAISE_7 = CW3_MALAISE_7, MALAISE_8 = CW3_MALAISE_8, MALAISE_9 = CW3_MALAISE_9, GAD2PHQ2_1 = CW3_GAD2PHQ2_1, GAD2PHQ2_2 = CW3_GAD2PHQ2_2, GAD2PHQ2_3 = CW3_GAD2PHQ2_3, GAD2PHQ2_4 = CW3_GAD2PHQ2_4)
long <- bind_rows(wave1_com, wave2_com, wave3_com)

```
## Changing labels and values, calculating mental health scores.
```{r}
str(long$COVID19)
long <- long %>%
  mutate(covid_positive = case_when(
      COVID19 == 1 ~ 1,
      COVID19 == 2 ~ 1,
      COVID19 %in% c(3, 4) ~ 0,
      TRUE ~ NA_real_)) %>%
  mutate(hospitalise = case_when(
    COVID_HOSPAD %in% NA ~ 0,
    COVID_HOSPAD == 2 ~ 0,
    COVID_HOSPAD == 1 ~ 1
  )) %>%
  rowwise() %>%
  mutate(malaise = sum(MALAISE_1, MALAISE_2, MALAISE_3, MALAISE_4, MALAISE_5, MALAISE_6, MALAISE_7, MALAISE_8, MALAISE_9, na.rm = FALSE)) %>%
  rowwise() %>%
  mutate(GAD = sum(GAD2PHQ2_1, GAD2PHQ2_2, GAD2PHQ2_3, GAD2PHQ2_4, na.rm = FALSE))
clean <- long %>%
  select(NCDSID, GHQ, HHNUM, wave, covid_positive, hospitalise, malaise, GAD)
```

