---
title: "How does COVID and loneliness affect mental health?"
output:
  html_document:
    df_print: paged
  pdf_document: default
date: "2025-05-06"
---

# Settings

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

# Reading data frames

```{r}
library(foreign)
library(tidyverse)
library(dplyr)
library(knitr)
library(lme4)      # For mixed-effects models
library(lmerTest)  # For p-values in mixed models
library(broom)     # For tidy model outputs
library(broom.mixed) # For mixed model outputs
library(corrr)  # For correlation analysis
library(mediation)
library(lavaan)       # For SEM approach (alternative)
library(knitr)
library(haven)
library("dagitty")
library("semPlot")
setwd("~/Desktop/LDAP_Project")
covid_19_wave1_survey_cls <- read_sav("covid-19_wave1_survey_cls.sav")
wave1 <- covid_19_wave1_survey_cls
covid_19_wave2_survey_cls <- read_sav("covid-19_wave2_survey_cls.sav")
wave2 <- covid_19_wave2_survey_cls
covid_19_wave3_survey_cls <- read_sav("covid-19_wave3_survey_cls.sav")
wave3 <- covid_19_wave3_survey_cls
```

# Picking variables

```{r}
wave1 <- wave1 %>%
  dplyr::select(NCDSID, CW1_COVID19, CW1_COVID_HOSPAD, CW1_GHQ, CW1_HHNUM, CW1_MALAISE_1, CW1_MALAISE_2, CW1_MALAISE_3, CW1_MALAISE_4, CW1_MALAISE_5, CW1_MALAISE_6, CW1_MALAISE_7, CW1_MALAISE_8, CW1_MALAISE_9, CW1_GAD2PHQ2_1, CW1_GAD2PHQ2_2, CW1_GAD2PHQ2_3, CW1_GAD2PHQ2_4, CW1_LONELY_1, CW1_LONELY_2, CW1_LONELY_3, CW1_LONELY_4)
wave2 <- wave2 %>%
  dplyr::select(NCDSID, CW2_COVID19, CW2_COVID_HOSPAD, CW2_GHQ, CW2_HHNUM, CW2_MALAISE_1, CW2_MALAISE_2, CW2_MALAISE_3, CW2_MALAISE_4, CW2_MALAISE_5, CW2_MALAISE_6, CW2_MALAISE_7, CW2_MALAISE_8, CW2_MALAISE_9, CW2_GAD2PHQ2_1, CW2_GAD2PHQ2_2, CW2_GAD2PHQ2_3, CW2_GAD2PHQ2_4, CW2_LONELY_1, CW2_LONELY_2, CW2_LONELY_3, CW2_LONELY_4)
wave3 <- wave3 %>%
  dplyr::select(NCDSID, CW3_COVID19, CW3_COVID_HOSPAD, CW3_GHQ, CW3_HHNUM, CW3_MALAISE_1, CW3_MALAISE_2, CW3_MALAISE_3, CW3_MALAISE_4, CW3_MALAISE_5, CW3_MALAISE_6, CW3_MALAISE_7, CW3_MALAISE_8, CW3_MALAISE_9, CW3_GAD2PHQ2_1, CW3_GAD2PHQ2_2, CW3_GAD2PHQ2_3, CW3_GAD2PHQ2_4, CW3_LONELY_1, CW3_LONELY_2, CW3_LONELY_3, CW3_LONELY_4)
```

# Data cleaning

```{r}
wave1 <- wave1 %>%
  filter(NCDSID != "")
wave2 <- wave2 %>%
  filter(NCDSID != "")
wave3 <- wave3 %>%
  filter(NCDSID != "")
```

# Data filtering

```{r}
common_ids <- reduce(list(wave1$NCDSID, wave2$NCDSID, wave3$NCDSID), intersect)
wave1_com <- wave1 %>%
  filter(NCDSID %in% common_ids)
wave2_com <- wave2 %>%
  filter(NCDSID %in% common_ids)
wave3_com <- wave3 %>%
  filter(NCDSID %in% common_ids)
```

# Specifying waves and long format

```{r}
wave1_com <- wave1_com %>%
  mutate(wave = 1)
wave2_com <- wave2_com %>%
  mutate(wave =2)
wave3_com <- wave3_com %>%
  mutate(wave = 3)

wave1_com <- wave1_com %>%
  rename(COVID19 = CW1_COVID19, COVID_HOSPAD = CW1_COVID_HOSPAD, GHQ = CW1_GHQ, HHNUM = CW1_HHNUM, MALAISE_1 = CW1_MALAISE_1, MALAISE_2 = CW1_MALAISE_2, MALAISE_3 = CW1_MALAISE_3, MALAISE_4 = CW1_MALAISE_4, MALAISE_5 = CW1_MALAISE_5, MALAISE_6 = CW1_MALAISE_6, MALAISE_7 = CW1_MALAISE_7, MALAISE_8 = CW1_MALAISE_8, MALAISE_9 = CW1_MALAISE_9, GAD2PHQ2_1 = CW1_GAD2PHQ2_1, GAD2PHQ2_2 = CW1_GAD2PHQ2_2, GAD2PHQ2_3 = CW1_GAD2PHQ2_3, GAD2PHQ2_4 = CW1_GAD2PHQ2_4, LONELY_1 = CW1_LONELY_1, LONELY_2 = CW1_LONELY_2, LONELY_3 = CW1_LONELY_3, LONELY_4 = CW1_LONELY_4)
wave2_com <- wave2_com %>%
  rename(COVID19 = CW2_COVID19, COVID_HOSPAD = CW2_COVID_HOSPAD, GHQ = CW2_GHQ, HHNUM = CW2_HHNUM, MALAISE_1 = CW2_MALAISE_1, MALAISE_2 = CW2_MALAISE_2, MALAISE_3 = CW2_MALAISE_3, MALAISE_4 = CW2_MALAISE_4, MALAISE_5 = CW2_MALAISE_5, MALAISE_6 = CW2_MALAISE_6, MALAISE_7 = CW2_MALAISE_7, MALAISE_8 = CW2_MALAISE_8, MALAISE_9 = CW2_MALAISE_9, GAD2PHQ2_1 = CW2_GAD2PHQ2_1, GAD2PHQ2_2 = CW2_GAD2PHQ2_2, GAD2PHQ2_3 = CW2_GAD2PHQ2_3, GAD2PHQ2_4 = CW2_GAD2PHQ2_4, LONELY_1 = CW2_LONELY_1, LONELY_2 = CW2_LONELY_2, LONELY_3 = CW2_LONELY_3, LONELY_4 = CW2_LONELY_4)
wave3_com <- wave3_com %>%
  rename(COVID19 = CW3_COVID19, COVID_HOSPAD = CW3_COVID_HOSPAD, GHQ = CW3_GHQ, HHNUM = CW3_HHNUM, MALAISE_1 = CW3_MALAISE_1, MALAISE_2 = CW3_MALAISE_2, MALAISE_3 = CW3_MALAISE_3, MALAISE_4 = CW3_MALAISE_4, MALAISE_5 = CW3_MALAISE_5, MALAISE_6 = CW3_MALAISE_6, MALAISE_7 = CW3_MALAISE_7, MALAISE_8 = CW3_MALAISE_8, MALAISE_9 = CW3_MALAISE_9, GAD2PHQ2_1 = CW3_GAD2PHQ2_1, GAD2PHQ2_2 = CW3_GAD2PHQ2_2, GAD2PHQ2_3 = CW3_GAD2PHQ2_3, GAD2PHQ2_4 = CW3_GAD2PHQ2_4, LONELY_1 = CW3_LONELY_1, LONELY_2 = CW3_LONELY_2, LONELY_3 = CW3_LONELY_3, LONELY_4 = CW3_LONELY_4)
long <- bind_rows(wave1_com, wave2_com, wave3_com)

```

# Changing labels, calculating values

```{r}
long <- long %>%
  mutate(covid_positive = case_when(
      COVID19 == 1 ~ 1,
      COVID19 == 2 ~ 1,
      COVID19 %in% c(3, 4) ~ 0,
      TRUE ~ NA_real_)) %>%
  mutate(hospitalise = case_when(
    COVID_HOSPAD %in% NA ~ 0,
    COVID_HOSPAD == 2 ~ 0,
    COVID_HOSPAD == 1 ~ 1
  )) %>%
  mutate(MALAISE_1 = case_when(
    MALAISE_1 == 2 ~ 0,
    MALAISE_1 == 1 ~ 1
  )) %>%
  mutate(MALAISE_2 = case_when(
    MALAISE_2 == 2 ~ 0,
    MALAISE_2 == 1 ~ 1
  )) %>%
  mutate(MALAISE_3 = case_when(
    MALAISE_3 == 2 ~ 0,
    MALAISE_3 == 1 ~ 1
  )) %>%
  mutate(MALAISE_4 = case_when(
    MALAISE_4 == 2 ~ 0,
    MALAISE_4 == 1 ~ 1
  )) %>%
  mutate(MALAISE_5 = case_when(
    MALAISE_5 == 2 ~ 0,
    MALAISE_5 == 1 ~ 1
  )) %>%
  mutate(MALAISE_6 = case_when(
    MALAISE_6 == 2 ~ 0,
    MALAISE_6 == 1 ~ 1
  )) %>%
  mutate(MALAISE_7 = case_when(
    MALAISE_7 == 2 ~ 0,
    MALAISE_7 == 1 ~ 1
  )) %>%
  mutate(MALAISE_8 = case_when(
    MALAISE_8 == 2 ~ 0,
    MALAISE_8 == 1 ~ 1
  )) %>%
  mutate(MALAISE_9 = case_when(
    MALAISE_9 == 2 ~ 0,
    MALAISE_9 == 1 ~ 1
  )) %>%
  rowwise() %>%
  mutate(malaise = sum(MALAISE_1, MALAISE_2, MALAISE_3, MALAISE_4, MALAISE_5, MALAISE_6, MALAISE_7, MALAISE_8, MALAISE_9, na.rm = FALSE)) %>%
  rowwise() %>%
  mutate(
    GAD = sum(GAD2PHQ2_1, GAD2PHQ2_2, GAD2PHQ2_3, GAD2PHQ2_4, na.rm = FALSE)) %>%
  rowwise() %>%
  mutate(
    LONELY = sum(LONELY_1, LONELY_2, LONELY_3, LONELY_4)
  )

```

```{r}
clean <- long %>%
  dplyr::select(NCDSID, GHQ, HHNUM, wave, covid_positive, hospitalise, malaise, GAD, LONELY)
# Finalised dataframe for multi-level analysis: clean
```

# Descriptive Stats
## 1. OVERALL SAMPLE CHARACTERISTICS
```{r}
cat("=== OVERALL SAMPLE CHARACTERISTICS ===\n")
cat("Total observations:", nrow(clean), "\n")
cat("Unique participants:", length(unique(clean$NCDSID)), "\n")
cat("Observations per wave:\n")
print(table(clean$wave))
```
## 2. DESCRIPTIVE STATISTICS BY WAVE
```{r}
cat("\n=== DESCRIPTIVE STATISTICS BY WAVE ===\n")

# Function to calculate descriptive stats
desc_stats <- function(x) {
  c(
    N = sum(!is.na(x)),
    Mean = round(mean(x, na.rm = TRUE), 2),
    SD = round(sd(x, na.rm = TRUE), 2),
    Min = min(x, na.rm = TRUE),
    Q1 = quantile(x, 0.25, na.rm = TRUE),
    Median = median(x, na.rm = TRUE),
    Q3 = quantile(x, 0.75, na.rm = TRUE),
    Max = max(x, na.rm = TRUE),
    Missing = sum(is.na(x))
  )
}

# Continuous variables by wave
continuous_vars <- c("GHQ", "malaise", "GAD", "LONELY")

desc_by_wave <- clean %>%
  group_by(wave) %>%
  summarise(
    across(all_of(continuous_vars), desc_stats, .names = "{.col}_{.fn}"),
    .groups = "drop"
  )

# Rename columns for better display
colnames(desc_by_wave) <- gsub("GHQ", "General Health Questionnaire", colnames(desc_by_wave))
colnames(desc_by_wave) <- gsub("malaise", "Malaise", colnames(desc_by_wave))
colnames(desc_by_wave) <- gsub("GAD", "Generalised Anxiety Disorder", colnames(desc_by_wave))
colnames(desc_by_wave) <- gsub("LONELY", "Loneliness Scale", colnames(desc_by_wave))


print(desc_by_wave)
```
## 3. CATEGORICAL VARIABLES BY WAVE
```{r}
cat("\n=== CATEGORICAL VARIABLES BY WAVE ===\n")

# COVID-19 positive cases by wave
cat("Covid Positive Cases by Wave:\n")
covid_table <- clean %>%
  group_by(wave, covid_positive) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = covid_positive, values_from = n, values_fill = 0) %>%
  mutate(
    Total = `0` + `1`,
    Percent_Positive = round((`1` / Total) * 100, 1)
  )
print(covid_table)

# Hospitalization by wave
cat("\nHospitalised by Wave:\n")
hosp_table <- clean %>%
  group_by(wave, hospitalise) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = hospitalise, values_from = n, values_fill = 0) %>%
  mutate(
    Total = `0` + `1`,
    Percent_Hospitalised = round((`1` / Total) * 100, 1)
  )
print(hosp_table)

# Household number distribution by wave
cat("\nHousehold Number Distribution by Wave:\n")
hhnum_summary <- clean %>%
  group_by(wave) %>%
  summarise(
    N = n(),
    Mean_HHNUM = round(mean(HHNUM, na.rm = TRUE), 2),
    SD_HHNUM = round(sd(HHNUM, na.rm = TRUE), 2),
    Min_HHNUM = min(HHNUM, na.rm = TRUE),
    Max_HHNUM = max(HHNUM, na.rm = TRUE),
    .groups = "drop"
  )
print(hhnum_summary)
```
## 4. DETAILED FREQUENCY TABLES
```{r}
cat("\n=== DETAILED FREQUENCY TABLES ===\n")

# GHQ scores by wave
cat("GHQ Score Distribution by Wave:\n")
ghq_freq <- clean %>%
  group_by(wave, GHQ) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(wave) %>%
  mutate(
    percent = round((n / sum(n)) * 100, 1),
    cum_percent = round(cumsum(percent), 1)
  )
print(ghq_freq)

# Malaise scores by wave (showing ranges to make it manageable)
cat("\nMalaise Score Ranges by Wave:\n")
malaise_ranges <- clean %>%
  mutate(
    malaise_range = case_when(
      malaise <= 2 ~ "0-2 (Low)",
      malaise <= 5 ~ "3-5 (Moderate)",
      malaise <= 8 ~ "6-8 (High)",
      malaise >= 9 ~ "9+ (Very High)",
      TRUE ~ "Missing"
    )
  ) %>%
  group_by(wave, malaise_range) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(wave) %>%
  mutate(percent = round((n / sum(n)) * 100, 1))
print(malaise_ranges)

# GAD scores by wave
cat("\nGAD Score Distribution by Wave:\n")
gad_freq <- clean %>%
  group_by(wave, GAD) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(wave) %>%
  mutate(
    percent = round((n / sum(n)) * 100, 1),
    cum_percent = round(cumsum(percent), 1)
  )
print(gad_freq)
```
## 5. CORRELATION MATRIX BY WAVE
```{r}
cat("\n=== CORRELATION MATRICES BY WAVE ===\n")

cor_matrices <- list()
for(w in 1:3) {
  wave_data <- clean %>% filter(wave == w)  # <-- FILTERS by wave!
  cor_vars <- c("GHQ", "malaise", "GAD", "LONELY", "covid_positive", "hospitalise")
  cor_matrix <- cor(wave_data[cor_vars], use = "pairwise.complete.obs")
  rownames(cor_matrix) <- colnames(cor_matrix) <- c(
    "General Health Questionnaire", "Malaise", 
    "Generalised Anxiety Disorder", "Loneliness Scale", 
    "Covid Positive", "Hospitalised"
  )
  cor_matrices[[paste0("Wave", w)]] <- cor_matrix
}


library(reshape2)

for(w in 1:3) {
  cor_matrix <- cor_matrices[[paste0("Wave", w)]]
  melted_cor <- melt(cor_matrix)
  
  cor_plot <- ggplot(melted_cor, aes(Var1, Var2, fill = value)) +
    geom_tile(color = "white") +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", 
                         midpoint = 0, limit = c(-1, 1), name = "r") +
    labs(
      title = paste("Correlation Matrix: Wave", w),
      x = NULL, y = NULL
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.title = element_text(hjust = 0.5)
    )
  print(cor_plot)}
```
## 6. MISSING DATA ANALYSIS
```{r}
cat("\n=== MISSING DATA ANALYSIS ===\n")
missing_analysis <- clean %>%
  group_by(wave) %>%
  summarise(
    N_total = n(),
    GHQ_missing = sum(is.na(GHQ)),
    malaise_missing = sum(is.na(malaise)),
    GAD_missing = sum(is.na(GAD)),
    LONELY_missing = sum(is.na(LONELY)),
    covid_positive_missing = sum(is.na(covid_positive)),
    hospitalise_missing = sum(is.na(hospitalise)),
    .groups = "drop"
  ) %>%
  mutate(
    across(ends_with("_missing"), ~ round((. / N_total) * 100, 1), .names = "{.col}_percent")
  )
print(missing_analysis)
```
## 7. SUMMARY TABLE FOR PUBLICATION
```{r}
cat("\n=== PUBLICATION-READY SUMMARY TABLE ===\n")
pub_table <- clean %>%
  group_by(wave) %>%
  summarise(
    N = n(),
    
    # Continuous variables: Mean (SD)
    GHQ_mean_sd = paste0(round(mean(GHQ, na.rm = TRUE), 2), " (", 
                         round(sd(GHQ, na.rm = TRUE), 2), ")"),
    
    malaise_mean_sd = paste0(round(mean(malaise, na.rm = TRUE), 2), " (", 
                            round(sd(malaise, na.rm = TRUE), 2), ")"),
    
    GAD_mean_sd = paste0(round(mean(GAD, na.rm = TRUE), 2), " (", 
                         round(sd(GAD, na.rm = TRUE), 2), ")"),
    
    LONELY_mean_sd = paste0(round(mean(LONELY, na.rm = TRUE), 2), " (", 
                            round(sd(LONELY, na.rm = TRUE), 2), ")"),
    
    # Categorical variables: n (%)
    COVID_positive = paste0(sum(covid_positive == 1, na.rm = TRUE), " (", 
                           round((sum(covid_positive == 1, na.rm = TRUE) / n()) * 100, 1), "%)"),
    
    Hospitalized = paste0(sum(hospitalise == 1, na.rm = TRUE), " (", 
                         round((sum(hospitalise == 1, na.rm = TRUE) / n()) * 100, 1), "%)"),
    
    .groups = "drop"
  )

print(pub_table)


  

cat("\n=== ANALYSIS COMPLETE ===\n")
cat("This comprehensive descriptive analysis provides:\n")
cat("1. Overall sample characteristics\n")
cat("2. Descriptive statistics by wave for continuous variables\n")
cat("3. Frequency distributions for categorical variables\n")
cat("4. Detailed frequency tables\n")
cat("5. Correlation matrices by wave\n")
cat("6. Missing data analysis\n")
cat("7. Publication-ready summary table\n")
```

## Visualising descriptive stats
```{r}
means_tbl <- clean %>%
  group_by(wave) %>%
  summarise(
    GHQ_mean = mean(GHQ, na.rm = TRUE),
    GHQ_sd = sd(GHQ, na.rm = TRUE),
    malaise_mean = mean(malaise, na.rm = TRUE),
    malaise_sd = sd(malaise, na.rm = TRUE),
    GAD_mean = mean(GAD, na.rm = TRUE),
    GAD_sd = sd(GAD, na.rm = TRUE),
    LONELY_mean = mean(LONELY, na.rm = TRUE),
    LONELY_sd = sd(LONELY, na.rm = TRUE)
  )
plot_tbl_long <- means_tbl %>%
  pivot_longer(
    cols = ends_with("_mean"),
    names_to = "Outcome",
    names_pattern = "(.*)_mean",
    values_to = "Mean"
  ) %>%
  mutate(SD = case_when(
    Outcome == "GHQ" ~ GHQ_sd,
    Outcome == "malaise" ~ malaise_sd,
    Outcome == "GAD" ~ GAD_sd,
    Outcome == "LONELY" ~ LONELY_sd
  ))
des_plot <- ggplot(plot_tbl_long, aes(x = as.factor(wave), y = Mean, group = Outcome, color = Outcome)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = Mean - SD, ymax = Mean + SD), width = 0.1) +
  labs(title = "Trends in Mental Health by Wave",
       x = "Wave", y = "Mean (±SD)") +
  theme_minimal()
des_plot
```


# Check the correlation / relationship between variables

## Hypothesis 1: Having COVID will predict poorer mental health
## Hypothesis 2: COVID severity predicts poorer mental health

## 1. DESCRIPTIVE COMPARISONS: COVID vs NO-COVID
```{r}
cat("=== COVID-19 AND MENTAL HEALTH PREDICTIVE ANALYSIS ===\n\n")

cat("1. DESCRIPTIVE COMPARISONS BY COVID STATUS\n")

# Create comparison table
covid_comparison <- clean %>%
  group_by(covid_positive) %>%
  summarise(
    N = n(),
    N_unique_participants = n_distinct(NCDSID),
    
    # Mental health outcomes
    `General Health Questionnaire` = paste0(round(mean(GHQ, na.rm = TRUE), 2), " (", 
                                           round(sd(GHQ, na.rm = TRUE), 2), ")"),
    
    `Malaise` = paste0(round(mean(malaise, na.rm = TRUE), 2), " (", 
                                     round(sd(malaise, na.rm = TRUE), 2), ")"),
    
    `Generalised Anxiety Disorder` = paste0(round(mean(GAD, na.rm = TRUE), 2), " (", 
                                           round(sd(GAD, na.rm = TRUE), 2), ")"),
    
    `Loneliness Scale` = paste0(round(mean(LONELY, na.rm = TRUE), 2), " (", 
                               round(sd(LONELY, na.rm = TRUE), 2), ")"),
    
    `Hospitalised Rate` = paste0(sum(hospitalise == 1, na.rm = TRUE), " (", 
                                round((sum(hospitalise == 1, na.rm = TRUE) / n()) * 100, 1), "%)"),
    
    .groups = "drop"
  ) %>%
  mutate(covid_positive = ifelse(covid_positive == 1, "Covid Positive", "Covid Negative"))

print(covid_comparison)



```
## 2. SIMPLE T-TESTS FOR EACH MENTAL HEALTH OUTCOME
```{r}
cat("\n\n2. T-TESTS: COVID POSITIVE vs COVID NEGATIVE\n")

mental_health_vars <- c("GHQ", "malaise", "GAD", "LONELY")
var_labels <- c("General Health Questionnaire", "Malaise", 
                "Generalised Anxiety Disorder", "Loneliness Scale")

ttest_results <- tibble()

for(i in 1:length(mental_health_vars)) {
  var <- mental_health_vars[i]
  label <- var_labels[i]
  
  # Perform t-test
  covid_pos <- clean %>% filter(covid_positive == 1) %>% pull(!!sym(var))
  covid_neg <- clean %>% filter(covid_positive == 0) %>% pull(!!sym(var))
  
  test_result <- t.test(covid_pos, covid_neg, na.rm = TRUE)
  
  # Calculate effect size (Cohen's d)
  pooled_sd <- sqrt(((length(covid_pos)-1)*var(covid_pos, na.rm=TRUE) + 
                     (length(covid_neg)-1)*var(covid_neg, na.rm=TRUE)) / 
                    (length(covid_pos) + length(covid_neg) - 2))
  cohens_d <- (mean(covid_pos, na.rm=TRUE) - mean(covid_neg, na.rm=TRUE)) / pooled_sd
  
  ttest_results <- bind_rows(ttest_results, 
                            tibble(
                              Variable = label,
                              COVID_Mean = round(mean(covid_pos, na.rm=TRUE), 2),
                              No_COVID_Mean = round(mean(covid_neg, na.rm=TRUE), 2),
                              Difference = round(COVID_Mean - No_COVID_Mean, 2),
                              t_statistic = round(test_result$statistic, 3),
                              p_value = round(test_result$p.value, 4),
                              Cohens_d = round(cohens_d, 3),
                              Effect_Size = case_when(
                                abs(cohens_d) < 0.2 ~ "Negligible",
                                abs(cohens_d) < 0.5 ~ "Small",
                                abs(cohens_d) < 0.8 ~ "Medium",
                                TRUE ~ "Large"
                              )
                            ))
}

print(ttest_results)
```
## 3. LONGITUDINAL MIXED-EFFECTS MODELS
```{r}
cat("\n\n3. MIXED-EFFECTS MODELS: LONGITUDINAL ANALYSIS\n")

# Create models for each mental health outcome
mixed_model_results <- tibble()

for(i in 1:length(mental_health_vars)) {
  var <- mental_health_vars[i]
  label <- var_labels[i]
  
  cat(paste("\nAnalyzing:", label, "\n"))
  
  # Basic model: COVID effect with random intercepts
  formula_basic <- as.formula(paste(var, "~ covid_positive + wave + (1|NCDSID)"))
  model_basic <- lmer(formula_basic, data = clean, REML = FALSE)
  
  # Enhanced model: Include hospitalization
  formula_enhanced <- as.formula(paste(var, "~ covid_positive + hospitalise + wave + (1|NCDSID)"))
  model_enhanced <- lmer(formula_enhanced, data = clean, REML = FALSE)
  
  # Extract results
  basic_summary <- tidy(model_basic, conf.int = TRUE)
  enhanced_summary <- tidy(model_enhanced, conf.int = TRUE)
  
  # Focus on COVID effect
  covid_effect_basic <- basic_summary %>% filter(term == "covid_positive")
  covid_effect_enhanced <- enhanced_summary %>% filter(term == "covid_positive")
  hosp_effect <- enhanced_summary %>% filter(term == "hospitalise")
  
  # Store results
  if(nrow(covid_effect_basic) > 0) {
    mixed_model_results <- bind_rows(mixed_model_results,
                                   tibble(
                                     Variable = label,
                                     Model = "Basic (COVID + Wave)",
                                     COVID_Coefficient = round(covid_effect_basic$estimate, 3),
                                     COVID_SE = round(covid_effect_basic$std.error, 3),
                                     COVID_p_value = round(covid_effect_basic$p.value, 4),
                                     COVID_CI_lower = round(covid_effect_basic$conf.low, 3),
                                     COVID_CI_upper = round(covid_effect_basic$conf.high, 3)
                                   ))
  }
  
  if(nrow(covid_effect_enhanced) > 0) {
    mixed_model_results <- bind_rows(mixed_model_results,
                                   tibble(
                                     Variable = label,
                                     Model = "Enhanced (+ Hospitalization)",
                                     COVID_Coefficient = round(covid_effect_enhanced$estimate, 3),
                                     COVID_SE = round(covid_effect_enhanced$std.error, 3),
                                     COVID_p_value = round(covid_effect_enhanced$p.value, 4),
                                     COVID_CI_lower = round(covid_effect_enhanced$conf.low, 3),
                                     COVID_CI_upper = round(covid_effect_enhanced$conf.high, 3)
                                   ))
  }
  
  # Print model comparison
  cat("Basic Model AIC:", round(AIC(model_basic), 2), "\n")
  cat("Enhanced Model AIC:", round(AIC(model_enhanced), 2), "\n")
  if(nrow(hosp_effect) > 0) {
    cat("Hospitalization Effect: β =", round(hosp_effect$estimate, 3), 
        ", p =", round(hosp_effect$p.value, 4), "\n")
  }
}

print(mixed_model_results)

mixed_model_results <- mixed_model_results %>%
  mutate(
    sig = case_when(
      COVID_p_value < 0.001 ~ "***",
      COVID_p_value < 0.01 ~ "**",
      COVID_p_value < 0.05 ~ "*",
      TRUE ~ ""
    )
  )
longi_plot <- ggplot(mixed_model_results, 
       aes(x = COVID_Coefficient, y = Variable, color = Model)) +
  geom_point(position = position_dodge(width = 0.6), size = 2) +
  geom_errorbarh(aes(xmin = COVID_CI_lower, xmax = COVID_CI_upper), 
                 height = 0.2, position = position_dodge(width = 0.6)) +
  geom_text(aes(label = sig), 
            position = position_dodge(width = 0.6), 
            hjust = -0.5, vjust = 0.5, size = 6) +  # adjusts text position and size
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  labs(title = "Effect of COVID on Mental Health Outcomes (Mixed-Effects Models)",
       x = "COVID Coefficient (β)", y = "Mental Health Outcome", color = "Model") +
  theme_minimal(base_size = 14)
longi_plot
```
## 4. TIMING ANALYSIS: WHEN DO EFFECTS APPEAR?
```{r}
cat("\n\n4. TIMING ANALYSIS: COVID EFFECTS BY WAVE\n")
cat("========================================\n")

# Analyze COVID effects within each wave
wave_analysis <- tibble()

for(wave_num in 1:3) {
  cat(paste("\nWave", wave_num, "Analysis:\n"))
  
  wave_data <- clean %>% filter(wave == wave_num)
  
  for(i in 1:length(mental_health_vars)) {
    var <- mental_health_vars[i]
    label <- var_labels[i]
    
    if(sum(!is.na(wave_data[[var]])) > 10) {  # Ensure sufficient data
      covid_pos <- wave_data %>% filter(covid_positive == 1) %>% pull(!!sym(var))
      covid_neg <- wave_data %>% filter(covid_positive == 0) %>% pull(!!sym(var))
      
      if(length(covid_pos) > 0 & length(covid_neg) > 0) {
        test_result <- t.test(covid_pos, covid_neg)
        
        wave_analysis <- bind_rows(wave_analysis,
                                 tibble(
                                   Wave = wave_num,
                                   Variable = label,
                                   COVID_Mean = round(mean(covid_pos, na.rm=TRUE), 2),
                                   No_COVID_Mean = round(mean(covid_neg, na.rm=TRUE), 2),
                                   Difference = round(COVID_Mean - No_COVID_Mean, 2),
                                   p_value = round(test_result$p.value, 4),
                                   Significant = ifelse(test_result$p.value < 0.05, "Yes", "No")
                                 ))
      }
    }
  }
}

print(wave_analysis)
```
## 5. HOSPITALIZATION AS SEVERITY INDICATOR
```{r}
cat("\n\n5. DOSE-RESPONSE ANALYSIS: COVID SEVERITY EFFECTS\n")

# Create COVID severity groups
clean_severity <- clean %>%
  mutate(
    covid_severity = case_when(
      covid_positive == 0 ~ "No COVID",
      covid_positive == 1 & hospitalise == 0 ~ "COVID - Not Hospitalized",
      covid_positive == 1 & hospitalise == 1 ~ "COVID - Hospitalized",
      TRUE ~ "Unknown"
    )
  ) %>%
  filter(covid_severity != "Unknown")

# ANOVA for each mental health variable
severity_results <- tibble()

for(i in 1:length(mental_health_vars)) {
  var <- mental_health_vars[i]
  label <- var_labels[i]
  
  # Perform ANOVA
  formula_anova <- as.formula(paste(var, "~ covid_severity"))
  anova_result <- aov(formula_anova, data = clean_severity)
  anova_summary <- summary(anova_result)
  
  # Post-hoc comparisons
  severity_means <- clean_severity %>%
    group_by(covid_severity) %>%
    summarise(
      N = n(),
      Mean = round(mean(!!sym(var), na.rm = TRUE), 2),
      SD = round(sd(!!sym(var), na.rm = TRUE), 2),
      .groups = "drop"
    )
  
  cat(paste("\n", label, "by COVID Severity:\n"))
  print(severity_means)
  
  f_statistic <- anova_summary[[1]][["F value"]][1]
  p_value <- anova_summary[[1]][["Pr(>F)"]][1]
  
  cat("F-statistic:", round(f_statistic, 3), ", p-value:", round(p_value, 4), "\n")
}
```
## 6. SUMMARY AND INTERPRETATION
```{r}
cat("\n\n6. SUMMARY AND INTERPRETATION\n")

cat("Key Findings:\n")
cat("1. Descriptive comparisons show mean differences between COVID+ and COVID- groups\n")
cat("2. T-tests provide statistical significance and effect sizes for group differences\n")
cat("3. Mixed-effects models account for repeated measures and individual differences\n")
cat("4. Wave-by-wave analysis shows when effects emerge or persist\n")
cat("5. Severity analysis tests dose-response relationship\n\n")

cat("Interpretation Guide:\n")
cat("- Positive coefficients = COVID associated with WORSE mental health\n")
cat("- Negative coefficients = COVID associated with BETTER mental health\n")
cat("- p < 0.05 = statistically significant effect\n")
cat("- Cohen's d: 0.2=small, 0.5=medium, 0.8=large effect\n\n")

cat("Next Steps:\n")
cat("1. Check model assumptions (residuals, normality)\n")
cat("2. Consider additional covariates (age, gender, SES)\n")
cat("3. Test for interaction effects (COVID × time)\n")
cat("4. Consider lagged effects (COVID in wave X predicting MH in wave X+1)\n")

cat("\n=== ANALYSIS COMPLETE ===\n")
```
# Hypothesis 3: Lonliness and hospitalisation (COVID severity) are mediators between the relationship of COVID positivity and mental health (general anxiety & malaise).
# SEM model with mediation
```{r}
cat("Running SEM analysis to explore direct effect (COVID positive on loneliness, anxiety and malaise.")
library(lavaan)

sem_model <- '
  # Mediation path
  LONELY ~ a * covid_positive
  
  # Outcomes with both direct and indirect paths
  GAD ~ b1 * LONELY + c1 * covid_positive
  malaise ~ b2 * LONELY + c2 * covid_positive

  # Indirect effects
  indirect_GAD := a * b1
  indirect_malaise := a * b2

  # Total effects
  total_GAD := c1 + (a * b1)
  total_malaise := c2 + (a * b2)
'

SEM_fit <- sem(sem_model, data = clean)
summary(SEM_fit, standardized = TRUE)

cat("The second SEM model adding in hospitalisation as a second predictor.")

sem_model_2 <- '
  # Predictors
  LONELY ~ a1 * covid_positive + a2 * hospitalise
  GAD ~ b1 * LONELY + c1 * covid_positive + c2 * hospitalise
  malaise ~ b2 * LONELY + d1 * covid_positive + d2 * hospitalise

  # Indirect effects
  indirect_GAD := a1 * b1
  indirect_GAD_hosp := a2 * b1
  indirect_malaise := a1 * b2
  indirect_malaise_hosp := a2 * b2

  # Total effects
  total_GAD := c1 + (a1 * b1) + c2 + (a2 * b1)
  total_malaise := d1 + (a1 * b2) + d2 + (a2 * b2)
'
SEM_fit_2 <- sem(sem_model_2, data = clean)
summary(SEM_fit_2, standardized = TRUE)
```
# Path diagram
```{r}
cat("The DAG diagram showing relationships between variables in SEM model 1")
dag <- dagitty("dag {
  covid_positive -> LONELY
  covid_positive -> GAD
  covid_positive -> malaise
  LONELY -> GAD
  LONELY -> malaise
}")
plot(dag)
cat("The SEM model1-based path diagram.")
semPaths(SEM_fit, whatLabels = "est", nodeLabels = c("Had COVID", "Loneliness", "GAD", "Malaise"), edge.label.cex = 0.8)

cat("The DAG diagram of SEM model 2 which added hospitalisation as a separate predictor.")

dag2 <- dagitty("dag {
  covid_positive -> hospitalise
  covid_positive -> LONELY
  covid_positive -> GAD
  covid_positive -> malaise
  hospitalise -> LONELY
  hospitalise -> GAD
  hospitalise -> malaise
  LONELY -> GAD
  LONELY -> malaise
}")
plot(dag2)

cat("The path diagram adding in Hospitalisation as a mediator.")
myLabels <- c("Had COVID", "Hospitalised", "Loneliness", "GAD", "Malaise")
SEM2 <- semPaths(SEM_fit_2, whatLabels = "est", nodeLabels = myLabels, edge.label.cex = 0.8)
SEM2


```
## Exploring longitudinal effects
```{r}
wide_data <- clean %>%
  pivot_wider(
    id_cols = NCDSID,
    names_from = wave,
    values_from = c(GAD, LONELY, malaise, covid_positive, hospitalise)
  )
longi_sem_model <- '
  # Autoregressive paths
  GAD_2 ~ GAD_1
  GAD_3 ~ GAD_2
  LONELY_2 ~ LONELY_1
  LONELY_3 ~ LONELY_2
  malaise_2 ~ malaise_1
  malaise_3 ~ malaise_2

  # Predictors at each wave
  GAD_1 ~ covid_positive_1 + hospitalise_1 + LONELY_1 + malaise_1
  GAD_2 ~ covid_positive_2 + hospitalise_2 + LONELY_2 + malaise_2
  GAD_3 ~ covid_positive_3 + hospitalise_3 + LONELY_3 + malaise_3

  malaise_1 ~ covid_positive_1 + hospitalise_1 + LONELY_1 + GAD_1
  malaise_2 ~ covid_positive_2 + hospitalise_2 + LONELY_2 + GAD_2
  malaise_3 ~ covid_positive_3 + hospitalise_3 + LONELY_3 + GAD_3

  LONELY_1 ~ covid_positive_1 + hospitalise_1
  LONELY_2 ~ covid_positive_2 + hospitalise_2
  LONELY_3 ~ covid_positive_3 + hospitalise_3

  # Cross-lagged effects (optional: if you want prior LONELY predicting later GAD)
  GAD_2 ~ LONELY_1
  GAD_3 ~ LONELY_2
  malaise_2 ~ LONELY_1
  malaise_3 ~ LONELY_2'

longi_fit <- sem(longi_sem_model, data = wide_data, missing = "fiml")
summary(longi_fit, standardized = TRUE)

```


## Generating a DAG diagram of longitudinal effect
```{r}

library(ggdag)

longi_dag <- dagitty('
dag {
  covid1 -> lonely1
  covid1 -> gad1
  covid1 -> malaise1
  hospital1 -> lonely1
  hospital1 -> gad1
  hospital1 -> malaise1
  lonely1 -> gad1
  lonely1 -> malaise1

  covid2 -> lonely2
  covid2 -> gad2
  covid2 -> malaise2
  hospital2 -> lonely2
  hospital2 -> gad2
  hospital2 -> malaise2
  lonely2 -> gad2
  lonely2 -> malaise2

  covid3 -> lonely3
  covid3 -> gad3
  covid3 -> malaise3
  hospital3 -> lonely3
  hospital3 -> gad3
  hospital3 -> malaise3
  lonely3 -> gad3
  lonely3 -> malaise3

  gad1 -> gad2
  gad2 -> gad3
  lonely1 -> lonely2
  lonely2 -> lonely3
  malaise1 -> malaise2
  malaise2 -> malaise3

  lonely1 -> gad2
  lonely2 -> gad3
  gad1 -> lonely2
  gad2 -> lonely3
}
')

ggdag(longi_dag, text = FALSE, use_labels = "name", layout = "tree")

```

## Generating path diagram for longitudinal SEM model.
```{r}
mod <- semPlot::semPlotModel(longi_fit)
mod@Vars$name
nodeLabels <- c(
  "GAD (W2)", "GAD (W3)", "Loneliness (W2)", "Loneliness (W3)",
  "Malaise (W2)", "Malaise (W3)", "GAD (W1)", "Malaise (W1)",
  "Loneliness (W1)", "COVID+ (W1)", "Hospitalised (W1)",
  "COVID+ (W2)", "Hospitalised (W2)", "COVID+ (W3)", "Hospitalised (W3)"
)
#I can't add this node labels in.
semPaths(longi_fit, whatLabels = "std", layout = "tree", border = FALSE, sizeMan = 5, edge.label.cex = 0.8)

```


